---
title: "Task2Košice"
author: "Anna Koliesnik"
date: "06/11/2025"
format:
  html:
    embed-resources: true
editor: source
---

## Question 1 Atomic Types
Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
# numeric values
distance_to_bratislava_km <- 400.0   # approx.
elevation_m <- 206                  # m (approx.)
area <- 242.8                       # km2 (approx Košice area)
avg_density <- 880                  # approx people/km2

# integer values
population <- 240000L
num_parks <- 25L
num_schools <- 50L

# character
name <- "Kosice"
old_name <- "Cassovia"
region <- "Košický kraj"
category <- "Mesto"

# logical
has_water <- TRUE        # Hornád river
is_inhabited <- TRUE
```


## Question 2 Vectors

Question description
Create three different vectors:

```{r}

# 1. Numeric vector
numeric_vector <- c(population, area)
numeric_vector

# 2. Sequence vector (from 20 to 100 step 5)
sequence_vector <- seq(from = 20, to = 100, by = 5)
sequence_vector

# 3. Logical vector of 10 random TRUE/FALSE 
set.seed(123)
logical_vector <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vector
```

## Question 3 Matrix

```{r}
set.seed(42)

m <- matrix(
  sample(-50:50, size = 9, replace = TRUE),
  nrow = 3, ncol = 3, byrow = TRUE
)

rownames(m) <- c("Stare_Mesto", "Terasa", "Nad_Jazerom")
colnames(m) <- c("Traffic", "Greenery", "Water")

m

row_avgs <- rowMeans(m)
col_avgs <- colMeans(m)

row_avgs
col_avgs
```


## Question 4 List

```{r}
kosicelist <- list(
  description = "Kosice mini-universe: basic facts, vectors for practice, and a city-themed matrix.",
  
  scalars = list(
    population = population,
    area = area,
    has_water = has_water,
    is_inhabited = is_inhabited,
    name = name,
    old_name = old_name,
    region = region,
    category = category
  ),
  
  vectors = list(
    numeric = numeric_vector,
    sequence = sequence_vector,
    logical = logical_vector
  ),
  
  matrix = m,
  
  matrix_stats = list(
    row_means = row_avgs,
    col_means = col_avgs
  )
)

str(kosicelist)
```


## Question 5 Factor

```{r}
development_level <- factor(
  c("medium", "high", "medium", "low", "high", "medium", "low"),
  levels = c("low", "medium", "high"), ordered = TRUE
)

table(development_level)
```


## Question 6 Data Frame

```{r}
kosice_df <- data.frame(
  district = c("Stare Mesto", "Terasa", "Nad Jazerom", "KVP", "Sever", "Barca"),
  pop = c(42000, 55000, 25000, 30000, 20000, 12000),
  has_playground = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
  zone = factor(
    c("mixed", "residential", "residential", "residential", "mixed", "industrial"),
    levels = c("residential", "mixed", "industrial", "recreational")
  ),
  dist_km = c(0.5, 3.0, 6.0, 4.5, 2.0, 7.0)
)

str(kosice_df)
summary(kosice_df)

kosice_df$priority <- with(
  kosice_df,
  ifelse(pop > 30000 & !has_playground, "high",
         ifelse(pop > 30000, "medium", "low"))
)

kosice_df$priority <- factor(kosice_df$priority, levels = c("low", "medium", "high"), ordered = TRUE)

table(kosice_df$zone)
table(kosice_df$priority)
kosice_df

write.csv(kosice_df, "kosice_df.csv", row.names = FALSE)
saveRDS(kosice_df, "kosice_df.rds")
```


```{r}
## --- Write to files ---
write.csv(kosice_df, "kosice_df.csv", row.names = FALSE)   # CSV
saveRDS(kosice_df, "kosice_df.rds")                        # keeps factors/attributes

```

## --- 2.1. Querying Vectors ---

```{r}
## 2.1.1 Sequence vector: 3rd element; elements 2:5; all > 20
third_elem <- sequence_vector[3]
third_elem

elem_2_to_5 <- sequence_vector[2:5]
elem_2_to_5

greater_than_20 <- sequence_vector[sequence_vector > 20]
greater_than_20
```

## 2.1.2 Logical vector: only TRUE values; indices of TRUE via which()

```{r}
only_true_values <- logical_vector[logical_vector]
only_true_values

true_indices <- which(logical_vector)
true_indices
```

## 2.2. Querying Your "Universe" List 

```{r}
## description via $
desc_text <- kosicelist$description
desc_text

## 3x3 matrix via [[ ]]
matrix_3x3 <- kosicelist[["matrix"]]
matrix_3x3

## second element from the numeric vector within the list
second_from_numeric_vec <- kosicelist$vectors$numeric[2]
second_from_numeric_vec
```

## 2.3. Investigating Your Data Frame 

```{r}
## first 3 rows
head_three <- head(kosice_df, 3)
head_three

## value at 4th row, 1st column
val_4_1 <- kosice_df[4, 1]
val_4_1

## logical column as a vector (has_playground)
playground_vec <- kosice_df$has_playground
playground_vec

## Logical subsetting: pop > 30000 AND has playground
df_subset <- kosice_df[kosice_df$pop > 30000 & kosice_df$has_playground, ]
df_subset

## %in% subsetting: keep only selected zones
df_subset_2 <- kosice_df[kosice_df$zone %in% c("residential", "mixed"), ]
df_subset_2
```

## 2.4. Handling "Corrupted" Data (NA) ---

```{r}
## copy original df
df_corrupted <- kosice_df

## introduce 3 NAs at specific locations
df_corrupted[2, "has_playground"] <- NA
df_corrupted[5, "dist_km"] <- NA
df_corrupted[6, "pop"] <- NA

## count NAs
total_NAs <- sum(is.na(df_corrupted))
total_NAs

NAs_by_col <- colSums(is.na(df_corrupted))
NAs_by_col

## mean with NA present (example: 'pop')
mean_pop_with_na <- mean(df_corrupted$pop, na.rm = TRUE)
mean_pop_with_na

## remove rows with any NA
df_clean <- na.omit(df_corrupted)
df_clean
```

## 2.5. Bonus: which.max / which.min 

```{r}
## index of the row with the highest population (original df)
highest_pop_index <- which.max(kosice_df$pop)
highest_pop_index

## print that row
kosice_df[highest_pop_index, ]

## also example for min distance
lowest_dist_index <- which.min(kosice_df$dist_km)
lowest_dist_index
kosice_df[lowest_dist_index, ]
```

